[Re] Non-additive coupling enables propagation of synchronous spiking activity in purely random networks

### 2018-03-13 article/Caze_2018.tex
Update manuscript and code for the new revision
Co-authored-by: Romain Caze <romain.caze@gmail.com>

\section*{Introduction}
Dendritic non-linearities increase neurons' computation capacity, turning them into complex computing units \cite{Poirazi2003a}. However network studies are usually based on point neuron models that do not incorporate dendrites and their non-linearities. In contrast, the study replicated here \cite{Memmesheimer2012} uses a simple point-neuron model that contains an effective description of dendrites by a non-linear summation of its excitatory synaptic input. Due to the simplicity of the model, both large-scale parameter exploration of a medium-sized network, as well as an analytical investigation of its properties are feasible.

@@

\section*{Methods}
\subsection*{Neural network model}
The simulations in the original paper were done in phase representation, but our implementation uses the more standard representation in terms of the neurons' membrane potential. The membrane potential dynamic $V_i$ (measured relative to the resting potential) of neuron $i$ is described by (a slightly rewritten version equivalent to equation (1) of \cite{Memmesheimer2012}):

\begin{equation}
\tau_m \frac{dV_i(t)}{dt} = - V_i(t) + V_0 + \sigma \big( \sum_{j\in E_i} w_\text{ex} \epsilon_{j}(t-\tau) \big) - \sum_{j \in I_i} w_\text{in}\epsilon_{j}(t-\tau),
\end{equation}
where $V(t)$ is the membrane voltage at time $t$, $\tau_m$ the membrane time constant, $V_0$ the displacement of the membrane potential due to a constant external input, $w_\text{ex}$ and $w_\text{in}$ respectively the excitatory and inhibitory weights, $\tau$ the transmission delay, and $E_i$ and $I_i$ the indices of neurons that connect with respectively excitatory and inhibitory weights to neuron $i$. If the membrane potential crosses a threshold $\Theta$, it emits the spike and its membrane potential is reset to its resting potential, i.e. $\SI{0}{\milli\volt}$. The function $\epsilon_{j}$ is 1 whenever neuron $j$ spikes, and 0 otherwise

The originality of this model lays in a filtering of the excitatory inputs by the function $\sigma$ to model dendritic non-linearities. The study considers two functions $\sigma$ (see Fig.~1 of the original paper and insets in Fig.~\ref{fig:markov}). Linear coupling, defined by $\sigma(x)=x$, and non-linear coupling with:
\[
    \sigma(x)= 
\begin{cases}
    x,              & \text{if } x\leq V_a\\
    V_a + \frac{V_c-V_a}{V_b-V_a} (x - V_a), & \text{if } V_a<x \leq V_b\\
    V_c,            & \text{otherwise}
\end{cases}
\]
For this non-linear function, the biological interpretation is that below a first threshold $V_a$, excitatory inputs are propagated passively by the dendrite and sum linearly. Above this threshold, they trigger a dendritic spike, resulting in an amplified somatic voltage. If they are even higher and cross the threshold $V_b$, they stop increasing the somatic voltage and saturate. Note that this non-linearity only affects spikes received synchronously as has been observed experimentally. With the introduction of this non-linearity, the modelled point neuron becomes an effective model of a neuron with one non-linear dendrite.

@@

\begin{table}[ht]
\begin{tabular}{ l  c  S p{2.85cm}}
  \toprule
  {Parameter} & {Symbol} & {Value} &\\
  \midrule
  Number of neurons & $N$ & 1000 &\\
  Connection probability & $p_0$ & 0.3 &\\
  Probability connection being excitatory & $p_\text{ex}$ & 0.5 &\\
  Excitatory connection weight & $w_\text{ex}$ & \SI{0.2}{\milli\volt} & (varied in Fig.~\ref{fig:grid})\\
  Inhibitory connection weight & $w_\text{in}$ & \SI{0.2}{\milli\volt} & (varied in Fig.~\ref{fig:grid})\\
  Membrane time constant & $ \tau_m $ & \SI{8}{\milli\second} &\\
  Firing threshold & $\Theta$ & \SI{16}{\milli\volt} &\\
  Synaptic delay & $\tau$ & \SI{5}{\milli\second} & (but see Implementation section)\\
  Threshold for supra-linear summation & $V_a$ & \SI{2}{\milli\volt} \\
  Threshold for saturation & $V_b$ & \SI{4}{\milli\volt} \\
  Saturation value & $V_c$ & \SI{6}{\milli\volt} \\
  \bottomrule
\end{tabular}
\caption{Parameter values used in the simulations}\label{tab:parameters}
\end{table}

@@

For non-linearly coupled networks, we have to use a slightly more complicated approach where we sum up all the received excitatory input in a variable \verb|ve|, apply the non-linear function $\sigma$ to this input, and then reset the variable \verb|ve| back to 0 to prepare for the next time step. In Brian syntax (the definition of the inhibitory synapses is identical to the linearly coupled network):
\begin{code}
eq = """dV/dt = -gamma*V + V0 : volt
        ve : volt
     """
G = br2.NeuronGroup(N, model=eq, threshold="V>Theta",
                    reset="V=0*mV", method='euler')
exc_syn = br2.Synapses(G, G, 'w : volt (constant, shared)',
                       on_pre='ve += w', delay=5 * ms - dt)
G.run_regularly('''
                V += clip(ve, 0*mV, 2*mV) + clip(2*(ve-2*mV), 0*mV, 4*mV)
                ve = 0*mV
                ''', when='after_synapses')
\end{code}
The \verb|run_regularly| operation that is run every time step to implement the non-linear $\sigma$ function refers to the predefined \verb|clip| function which clips the value to a given range, following the syntax
\begin{code}
clip(value, lower_bound, upper_bound)
\end{code}
To facilitate simulation, we use the \emph{joblib} library to automatically store results and to distribute computation over processor cores. Several simulation and analysis functions are annotated with a \verb|@mem.cache| decorator, meaning that their input arguments and outputs will be automatically stored to disk and calling the function will return the stored values if available. This feature does not only ease development (e.g. plotting functions can be changed without re-running the underlying simulations), but also allows to interrupt long-running parameter explorations without losing the results obtained so far. The same result could be achieved by manually storing results to disk, but joblib's caching mechanism helps to avoid common errors (e.g. to re-use an old result obtained with an outdated version of the code) while keeping the code readable. To reduce the total simulation time, the grid exploration in Fig.~\ref{fig:grid}, and the analysis of group size evolution in Fig.~\ref{fig:markov} have been distributed over processor cores using joblib's \verb|Parallel| construct. For the simulations presented in Fig.~\ref{fig:grid}, multiple repetitions for the same combination of synaptic weights are distributed over processor cores, for the simulations presented in Fig.~\ref{fig:markov}, simulations are evaluated in parallel over group sizes. In both cases, a number of $C-1$ parallel processes is used, where $C$ is the number of processor cores in the user's system.

@@

For the grid search presented in Fig.~\ref{fig:grid}, we varied inhibitory and excitatory synaptic weights between \SI{0.16}{\milli\volt} and \SI{0.4}{\milli\volt} (given that each neuron receives on average 150 excitatory and 150 inhibitory inputs, this corresponds to a ``total weight'' between \SI{24}{\milli\volt} and \SI{60}{\milli\volt}, as used on the axes of Fig.~\ref{fig:grid}), using 100 steps for each variable (comparable to the 96 steps used in the original publication). Each network was simulated 10 times with different initial conditions (synaptic connections and initial membrane potential values), whereas the original publication used 20 simulations per weight combination and additionally added ``1--50 random spikes initially in transit''.

To numerically derive the evolution of synchronous pulse sizes presented in Fig.~\ref{fig:markov} (for the semi-analytical approach, see the next section), we performed simulations of \SI{56}{\milli\second} each. During each simulation, we stimulated the network at 50ms with a synchronous pulse of activity, consisting of between 1 and 176 neurons (varied in steps of 5 -- the original study used group sizes between 1 and 181 varied in steps of 6). We then observed the number of active neurons at \SI{55}{\milli\second}, i.e. at one $\tau$ after the stimulation. For each stimulation group size, we ran the simulations 50 times with different initial conditions, consistent with the original study.

\subsection*{Semi-analytical and analytical methods}
We did not attempt to replicate the analytical solution for the evolution of synchronous pulses based on diffusion approximation (blue dots in Fig.~4 of \cite{Memmesheimer2012}).
To calculate the semi-analytical solutions for Fig.~\ref{fig:markov}, based on a numerical estimate of the distribution $P(V)$, we proceeded as follows. First, we ran $2\times 50$ (once for the linear, once for the non-linear network) simulations of \SI{250}{\milli\second} each of a network with the default parameters but without any additional synchronous stimulation.  We binned all membrane potential values over all time steps and neurons into 100 bins in the voltage interval $[-\Theta/8, \Theta]$, to obtain an estimate of $P(V)$, the distribution of the membrane potential. From this distribution we can estimate the probability $F$ that an additional input $\epsilon$ drives the neuron over its threshold (or equivalently, that the neuron is less than $\epsilon$ away from its threshold):
\begin{equation}
F(\epsilon) = P(V>\Theta-\epsilon) = \int_{\Theta-\epsilon}^{\Theta}P(V)dV
\end{equation}
Let $\epsilon(n_\text{ex}, n_\text{in})$ be the total input to a neuron that receives $n_\text{ex}$ excitatory and $n_\text{in}$ inhibitory spikes, i.e. $\epsilon(n_\text{ex}, n_\text{in}) = \sigma(n_\text{ex}w_\text{ex}) + n_\text{in}w_\text{in}$. Let us assume that in the last iteration $g_i$ neurons spiked and that these neurons are refractory now, i.e. that $N-g_i$ neurons are available to spike in this iteration. To calculate the probability for a single (non-refractory) neuron $j$ to spike in response to these $g_i$ neurons, we will first calculate the probability that out of the $g_i$ neurons that spiked, exactly $n_\text{ex}$ neurons are connected with excitatory connections and $n_\text{in}$ neurons are connected with inhibitory connections to neuron $j$. There are $\binom{g_i}{n_\text{ex} + n_\text{in}}$ possibilities to chose $n_\text{ex} + n_\text{in}$ out of $g_i$ neurons, and $\binom{n_\text{ex} + n_\text{in}}{n_\text{ex}}$ possibilities to divide these neurons into groups of size $n_\text{ex}$ and $n_\text{in}$. This yields the total number of combinations $c(g_i, n_\text{ex}, n_\text{in})$:
\begin{equation}
c(g_i, n_\text{ex}, n_\text{in}) = \binom{g_i}{n_\text{ex} + n_\text{in}}\binom{n_\text{ex} + n_\text{in}}{n_\text{ex}} = \frac{g_i!}{n_\text{ex}!n_\text{in}!(g_i - n_\text{ex} - n_\text{in})!}
\end{equation}
The probability that these selected $n_\text{ex}$ neurons are actually connected via excitatory connections to neuron $j$, that the $n_\text{in}$ neurons are actually connected via inhibitory connections to neuron $j$, and that the remaining $g_i - n_\text{ex} - n_\text{in}$ neurons are \emph{not} connected to neuron $j$ is:
\begin{equation}
p(g_i, n_\text{ex}, n_\text{in}) = (p_0 p_\text{ex})^{n_\text{ex}} (p_0 p_\text{in})^{n_\text{in}} (1 - p_0)^{g_i - n_\text{ex} - n_\text{in}}
\end{equation}
Finally, we need to sum these probabilities up over all possible values for $n_\text{ex}$ and $n_\text{in}$, multiplying it with the probability that the generated input elicits a spike. This yiels the total probability $P^s(g_i)$ that a neuron $j$ spikes in response to the activity of $g_i$ neurons in the previous iteration:
\begin{equation}
P^s(g_i) = \sum_{n_\text{ex}=1}^{g_i}\sum_{n_\text{in}=0}^{g_i-n_\text{ex}} F\big(\epsilon(n_\text{ex}, n_\text{in})\big) c(g_i, n_\text{ex}, n_\text{in}) p(g_i, n_\text{ex}, n_\text{in})
\end{equation}
In Fig.~\ref{fig:markov}, we plot the expected value of spiking neurons during the next iteration, i.e.:
\begin{equation}
E(g_{i+1}|g_i = g_i') = (N - g_i') P^s(g_i')
\end{equation}

@@

To confirm the equivalence of our network's behaviour with the original study over a wide range of synaptic connection strength we ran a grid search exploration for linearly (Fig.~\ref{fig:grid}A) and non-linearly (Fig.~\ref{fig:grid}B) coupled networks. Network behaviour is colour-coded as being part of one out of four classes: stable activity, but no persistent propagation (green); stable activity with persistent propagation (blue); unstable activity after synchronous stimulation (yellow); unstable activity before synchronous stimulation (red). For quantitative definitions, see Table~\ref{tab:color}. We run each combination of connection strengths 10 times, and combine the results to yield a color in RGB notation (values between 0 and 1 for the red, green, and blue component), according to:
\begin{equation}
R = U_1 + U_2;\quad G = E + U_2;\quad B = S,
\end{equation}
where $U_1$, and $U_2$ stand for the fraction of trials displaying unstable network activity before respectively after onset of propagation, and $S$ and $E$ for the fraction of trials displaying stable background activity with respectively without persistent propagation. We obtained this scheme directly from the authors of the original study, it was only coarsely described in the original publication. Combining individual results in this way does not necessarily lead to easily interpretable colours, but using a different scheme, e.g. colouring according to the class that occurred the most often, leads to visually similar results (not shown).

Our replication does not reproduce all details of the original plots (in particular at the top of the plots), but looks very similar on a coarse scale. Importantly, it strongly supports the original paper's main conclusion that non-linearly coupled networks have a large parameter range of stable propagation (i.e., blue-coloured points), while linearly coupled networks do not. The minor differences in our replication could be due to either our use of a clock-based simulation method, or to the slightly different initial conditions (no spikes ``in transit'').
\begin{table}
\begin{tabular}{ c  l  p{0.32\textwidth}}
  \toprule
   & Class & Definition \\
  \midrule
  $U_1$ (red) & Unstable before stimulation & $\exists t<t_\text{stim}: a_b(t) \geq 100$ \\
  $U_2$ (yellow) & Unstable after stimulation & $\exists t>t_\text{stim}: a_b(t) \geq 100$ \\
  $E$ (green) &  Stable activity, no propagation & $\forall t: a_b(t) < 100\, \wedge\, g_\text{p} < 10$\\
  $S$ (blue) & Stable activity, persistent propagation & $\forall t: a_b(t) < 100\, \wedge\, g_\text{p} \geq 10$\\
  \bottomrule
\end{tabular}
\caption{Classification of network behaviour for Fig.~\ref{fig:grid}. Here, $a_b$ refers to the background activity, the number of neurons active during a \SI{1}{\milli\second} time bin, ignoring the time steps at $t_\text{p}^k = t_\text{stim} + k\cdot\tau$, i.e. time steps that follow the synchronous stimulation by multiples of the synaptic delay. The variable $g_\text{p}$ refers to the number of successful propagations of the synchronous activity, i.e. the number of time steps $t_\text{p}^k$ where the number of active neurons exceeds the maximal background activity $a_b$.}\label{tab:color}
\end{table}

\begin{figure}
\centering
\begin{subfigure}[t]{0.5\textwidth}
\caption{}\vspace{-1ex}
\includegraphics[width=\textwidth]{Figs/grid_l_dt_100us_mean.pdf}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\caption{}\vspace{-1ex}
\includegraphics[width=\textwidth]{Figs/grid_nl_dt_100us_mean.pdf}
\end{subfigure}
\caption{\label{fig:grid}
\textbf{Qualitative network behaviour as a function of excitatory and inhibitory connection strength for linearly (A) and non-linearly (B) coupled networks.} For each combination, we initiated synchronous activity with 100 neurons and assessed the stability of the temporal evolution. Blue colouring indicates stable propagation of synchrony, red indicates unstable background activity before onset of propagation, yellow indicates unstable background activity after onset of propagation, and green colouring indicates the absence of propagation (see Methods for details).\\
Replication of Fig.~3a and b in \cite{Memmesheimer2012} (\url{https://doi.org/10.1371/journal.pcbi.1002384.g003})}
\end{figure}

>>>>

\section*{Introduction}
Dendritic non-linearities increase neurons' computation capacity, turning them into complex computing units \cite{Poirazi2003a}. However, network studies are usually based on point neuron models that do not incorporate dendrites and their non-linearities. In contrast, the study replicated here \cite{Memmesheimer2012} uses a simple point-neuron model that contains an effective description of dendrites by a non-linear summation of its excitatory synaptic input. Due to the simplicity of the model, both large-scale parameter exploration of a medium-sized network, as well as an analytical investigation of its properties are feasible.

@@

\section*{Methods}
\subsection*{Neural network model}
The simulations in the original paper were done in phase representation, but our implementation uses the more standard representation in terms of the neurons' membrane potential. The membrane potential $V_i$ of neuron $i$, measured relative to the resting potential, is described by:

\begin{equation}
\label{eq:neuron_model}
\tau_m \frac{dV_i(t)}{dt} = - V_i(t) + V_0 + \sigma \left( \sum_{j\in E_i} w_\text{ex} \epsilon_{j}(t-\tau) \right) - \sum_{j \in I_i} w_\text{in}\epsilon_{j}(t-\tau),
\end{equation}
where $V(t)$ is the membrane voltage at time $t$, $\tau_m$ the membrane time constant, $V_0$ the displacement of the membrane potential due to a constant external input, $w_\text{ex}$ and $w_\text{in}$ respectively the excitatory and inhibitory weights, $\tau$ the transmission delay, and $E_i$ and $I_i$ the indices of neurons that connect with respectively excitatory and inhibitory weights to neuron $i$. If the membrane potential crosses a threshold $\Theta$, it emits the spike and its membrane potential is reset to its resting potential, i.e. $\SI{0}{\milli\volt}$. The function $\epsilon_{j}$ is 1 whenever neuron $j$ spikes, and 0 otherwise. Equation~\ref{eq:neuron_model} is an alternative, mathematically equivalent formulation of equation (1) of \cite{Memmesheimer2012}.

The originality of this model lays in a filtering of the excitatory inputs by the function $\sigma$ to model dendritic non-linearities. The study considers two functions $\sigma$ (see Fig.~1 of the original paper and insets in Fig.~\ref{fig:markov_corrected}). Linear coupling, defined by $\sigma(x)=x$, and non-linear coupling with:
\begin{equation}
    \sigma(x)= 
\begin{cases}
    x              & \text{, if } x\leq V_a\\
    V_a + \frac{V_c-V_a}{V_b-V_a} (x - V_a), & \text{, if } V_a<x \leq V_b\\
    V_c            & \text{, otherwise.}
\end{cases}
\label{eq:non-linear}
\end{equation}
For this non-linear function, the biological interpretation is that below a first threshold $V_a$, excitatory inputs are propagated passively by the dendrite and sum linearly. Above this threshold, they trigger a dendritic spike, resulting in an amplified somatic voltage. If they are even higher and cross the threshold $V_b$, they stop increasing the somatic voltage and saturate. Note that this non-linearity only affects spikes received synchronously as has been observed experimentally \cite{Softky1994}. With the introduction of this non-linearity, the modelled point neuron becomes an effective model of a neuron with one non-linear dendrite.

@@

\begin{table}[ht]
\begin{tabular}{ l  c  S p{2.85cm}}
  \toprule
  {Parameter} & {Symbol} & {Value} &\\
  \midrule
  Number of neurons & $N$ & 1000 &\\
  Connection probability & $p_0$ & 0.3 &\\
  Probability connection being excitatory & $p_\text{ex}$ & 0.5 &\\
  Excitatory connection weight & $w_\text{ex}$ & \SI{0.2}{\milli\volt} & (varied in Fig.~\ref{fig:grid})\\
  Inhibitory connection weight & $w_\text{in}$ & \SI{0.2}{\milli\volt} & (varied in Fig.~\ref{fig:grid})\\
  Membrane time constant & $ \tau_m $ & \SI{8}{\milli\second} &\\
  Firing threshold & $\Theta$ & \SI{16}{\milli\volt} &\\
  External input & $V_0$ & \SI{17.6}{\milli\volt}&(effect on $V$)\\
  Synaptic delay & $\tau$ & \SI{5}{\milli\second} & (but see Implementation section)\\
  Threshold for supra-linear summation & $V_a$ & \SI{2}{\milli\volt} \\
  Threshold for saturation & $V_b$ & \SI{4}{\milli\volt} \\
  Saturation value & $V_c$ & \SI{6}{\milli\volt} \\
  \bottomrule
\end{tabular}
\caption{Parameter values used in the simulations}\label{tab:parameters}
\end{table}

@@

We used a the predefined \verb|clip| function to implemented non-linear coupling into a network. The function clips a value to a given range, following the syntax
\begin{code}
clip(value, lower_bound, upper_bound)
\end{code}

We employed this functions on the received excitatory inputs stored in a placed holder variable \verb|ve|. The variable corresponds to x in eq.~\ref{eq:non-linear}. We then reset the variable \verb|ve| back to 0 to prepare for the next time step. In Brian syntax (the definition of the inhibitory synapses is identical to the linearly coupled network):

\begin{code}
eq = """dV/dt = -gamma*V + V0 : volt
        ve : volt
     """
G = br2.NeuronGroup(N, model=eq, threshold="V>Theta",
                    reset="V=0*mV", method='euler')
exc_syn = br2.Synapses(G, G, 'w : volt (constant, shared)',
                       on_pre='ve += w', delay=5 * ms - dt)
G.run_regularly('''
                V += clip(ve, 0*mV, 2*mV) + clip(2*(ve-2*mV), 0*mV, 4*mV)
                ve = 0*mV
                ''', when='after_synapses')
\end{code}

The \verb|run_regularly| operation applies the non-linearity to the summed input at every time step.

In addition to random initial conditions, i.e. random initial values of the membrane potential, the simulation includes 1--50 random spikes ``initially in transit''. The original article does not go into further details about their implementation, so we decided to use the following: at the beginning of the simulation, a uniformly distributed number between 1 and 50 is drawn to determine the total number of initial spikes. These spikes are ``in transit'', meaning that they are modelled as arising from events before the start of the simulation. Since the synaptic delay in the network is \SI{5}{\milli\second}, we draw the arrival times for these spikes between \SI{0}{\milli\second} and \SI{5}{\milli\second}. Consistent with the probability for excitatory connections in the network ($p_\text{ex}=0.5$), we take half of these spikes as arising from excitatory and half from inhibitory sources, modelling their effect as described above.

To facilitate simulation, we use the \emph{joblib} library to automatically store results and to distribute computation over processor cores. Several simulation and analysis functions are annotated with a \verb|@mem.cache| decorator, meaning that their input arguments and outputs will be automatically stored to disk and calling the function will return the stored values if available. This feature does not only ease development (e.g. plotting functions can be changed without re-running the underlying simulations), but also allows to interrupt long-running parameter explorations without losing the results obtained so far. The same result could be achieved by manually storing results to disk, but joblib's caching mechanism helps to avoid common errors (e.g. to re-use an old result obtained with an outdated version of the code) while keeping the code readable. To reduce the total simulation time, the grid exploration in Fig.~\ref{fig:grid}, and the analysis of group size evolution in Fig.~\ref{fig:markov_corrected} have been distributed over processor cores using joblib's \verb|Parallel| construct. For the simulations presented in Fig.~\ref{fig:grid}, multiple repetitions for the same combination of synaptic weights are distributed over processor cores, for the simulations presented in Fig.~\ref{fig:markov_corrected}, simulations are evaluated in parallel over group sizes. In both cases, a number of $C-1$ parallel processes is used, where $C$ is the number of processor cores in the user's system.

@@

For the grid search presented in Fig.~\ref{fig:grid}, we varied inhibitory and excitatory synaptic weights between \SI{0.16}{\milli\volt} and \SI{0.4}{\milli\volt}. Given that each neuron receives on average 150 excitatory and 150 inhibitory inputs, this corresponds to a ``total weight'' between \SI{24}{\milli\volt} and \SI{60}{\milli\volt}, as used on the axes of Fig.~\ref{fig:grid}. We used a resolution of 100 steps for each variable, comparable to the 96 steps used in the original publication. Each network was simulated 10 times with different initial conditions (synaptic connections and initial membrane potential values), whereas the original publication used 20 simulations per weight combination.

To numerically derive the evolution of synchronous pulse sizes presented in Fig.~\ref{fig:markov_corrected} (for the semi-analytical approach, see the next section), we performed simulations of \SI{56}{\milli\second} each. During each simulation, we stimulated the network at \SI{50}{\milli\second} with a synchronous pulse of activity, consisting of between 1 and 181 neurons (varied in steps of 5 -- the original study used group sizes in the same range varied in steps of 6). We then observed the number of active neurons at \SI{55}{\milli\second}, i.e. at one $\tau$ after the stimulation. For each stimulation group size, we ran the simulations 50~times with different initial conditions, consistent with the original study. To achieve a better quantitative match with the original study, we applied a correction for the stimulation group size $g'_0$ and the observed group size $g'_1$ in the next iteration (for the uncorrected results, see Fig.~\ref{fig:markov}). We refer to the corrected values as $\hat{g}_0$ and $\hat{g}_1$, respectively. These corrections are necessary because of the background activity and our clock-based simulation method and are motivated as follows.  The activity of the full network in the absence of stimulation is about \SI{55}{\kilo\hertz} (see Fig.~\ref{fig:raster}). Given that our simulation uses a time step of \SI{0.1}{\milli\second}, this means that at every time step about 5.5 neurons on average are synchronously active. When we stimulate a network with a pulse of $g'_0$ synchronous neurons, the actual number of synchronously active neurons is therefore higher than this stimulation. Similarly, we have to correct the size of the observed group in the next iteration. In contrast to the simulation strategy of the original publication that could make use of an ``infinite precision'' to detect synchronous spikes triggered by the stimulation, our simulation will also detect all the synchronous spikes resulting from the background activity. In the corrected values shown in Fig.~\ref{fig:markov_corrected} we therefore used the actually observed synchronous spikes $\hat{g}'_0 \geq g'_0$ at the time of the stimulation, and the number of added spikes due to the stimulation $\hat{g}'_1 = g'_{1 \text{stim}} - g'_{1 \text{no stim}}$, where $g'_{1 \text{stim}}$ is the number of synchronous spikes we observe in a network with stimulation, and $g'_{1 \text{no stim}}$ the number of observed spikes in a network without stimulation. Both networks are simulated with the same random seed and therefore have the same network connectivity and the same initial conditions; the only difference lies in the presence or absence of external synchronous stimulation. Note that the correction for $\hat{g}'_1$ slightly ``overcorrects'', as it discards all spikes that are resulting from background activity, even those that also would have been triggered by the stimulation.

\subsection*{Semi-analytical and analytical methods}
We did not attempt to replicate the analytical solution for the evolution of synchronous pulses based on diffusion approximation (blue dots in Fig.~4 of \cite{Memmesheimer2012}), since the original publication did not give a detailed description of this approach. However, we replicate the semi-analytical work that was described in more detail.

To calculate the semi-analytical solutions for Fig.~\ref{fig:markov_corrected}, based on a numerical estimate of the distribution $P(V)$, we proceeded as follows. First, we ran $2\times 50$ (once for the linear, once for the non-linear network) simulations of \SI{250}{\milli\second}. Each of these simulations was randomly connected and had random initial conditions, but received no additional synchronous stimulation. The original study additionally simulated each of the 100 networks 10 times, with the same synaptic connectivity but differing in the initial conditions. We then binned all membrane potential values over all time steps and neurons into 100 bins in the voltage interval $[-\Theta/8, \Theta]$ to obtain an estimate of $P(V)$, the distribution of the membrane potential. Note that the original study did not further specify how $P(V)$ was estimated from the simulation results.

We then calculate the expected synchronous group size after one iteration as in the original publication. To make this calculation clearer, we now describe its derivation in detail. From the distribution $P(V)$ we can estimate the probability $F$ that an additional input $\epsilon$ drives the neuron over its threshold (or equivalently, that the neuron is less than $\epsilon$ away from its threshold):
\begin{equation}
F(\epsilon) = P(V>\Theta-\epsilon) = \int_{\Theta-\epsilon}^{\Theta}P(V)dV
\end{equation}
Let $\epsilon(n_\text{ex}, n_\text{in})$ be the total input to a neuron that receives $n_\text{ex}$ excitatory and $n_\text{in}$ inhibitory spikes, i.e. $\epsilon(n_\text{ex}, n_\text{in}) = \sigma(n_\text{ex}w_\text{ex}) + n_\text{in}w_\text{in}$. Let us assume that in the previous iteration $g_i$ neurons spiked and that these neurons are refractory now, i.e. that $N-g_i$ neurons are available to spike in this iteration. To calculate the probability for a single (non-refractory) neuron $j$ to spike in response to these $g_i$ neurons, we will first calculate the probability that out of the $g_i$ neurons that spiked, exactly $n_\text{ex}$ neurons are connected with excitatory connections and $n_\text{in}$ neurons are connected with inhibitory connections to neuron $j$. There are $\binom{g_i}{n_\text{ex} + n_\text{in}}$ possibilities to chose $n_\text{ex} + n_\text{in}$ out of $g_i$ neurons, and $\binom{n_\text{ex} + n_\text{in}}{n_\text{ex}}$ possibilities to divide these neurons into groups of size $n_\text{ex}$ and $n_\text{in}$. This yields the total number of combinations $c(g_i, n_\text{ex}, n_\text{in})$:
\begin{equation}
c(g_i, n_\text{ex}, n_\text{in}) = \binom{g_i}{n_\text{ex} + n_\text{in}}\binom{n_\text{ex} + n_\text{in}}{n_\text{ex}} = \frac{g_i!}{n_\text{ex}!n_\text{in}!(g_i - n_\text{ex} - n_\text{in})!}
\end{equation}
The probability that these selected $n_\text{ex}$ neurons are actually connected via excitatory connections to neuron $j$, that the $n_\text{in}$ neurons are actually connected via inhibitory connections to neuron $j$, and that the remaining $g_i - n_\text{ex} - n_\text{in}$ neurons are \emph{not} connected to neuron $j$ is:
\begin{equation}
p(g_i, n_\text{ex}, n_\text{in}) = (p_0 p_\text{ex})^{n_\text{ex}} (p_0 p_\text{in})^{n_\text{in}} (1 - p_0)^{g_i - n_\text{ex} - n_\text{in}}
\end{equation}
Finally, we need to sum these probabilities up over all possible values for $n_\text{ex}$ and $n_\text{in}$, multiplying it with the probability that the generated input elicits a spike. This yiels the total probability $P^s(g_i)$ that a neuron $j$ spikes in response to the activity of $g_i$ neurons in the previous iteration:
\begin{equation}
P^s(g_i) = \sum_{n_\text{ex}=1}^{g_i}\sum_{n_\text{in}=0}^{g_i-n_\text{ex}} F\big(\epsilon(n_\text{ex}, n_\text{in})\big) c(g_i, n_\text{ex}, n_\text{in}) p(g_i, n_\text{ex}, n_\text{in})
\end{equation}
In Fig.~\ref{fig:markov_corrected}, we plot the expected value of spiking neurons during the next iteration, given a certain initial synchronous pulse $g_i'$:
\begin{equation}
E(g_{i+1}|g_i = g_i') = (N - g_i') P^s(g_i')
\end{equation}

@@

To confirm the equivalence of our network's behaviour with the original study over a wide range of synaptic connection strength we ran a grid search exploration for linearly (Fig.~\ref{fig:grid}A) and non-linearly (Fig.~\ref{fig:grid}B) coupled networks. Network behaviour is colour-coded as being part of one out of four classes: stable activity, but no persistent propagation (green); stable activity with persistent propagation (blue); unstable activity after synchronous stimulation (yellow); unstable activity before synchronous stimulation (red). For quantitative definitions, see Table~\ref{tab:color}. We run each combination of connection strengths 10 times, and combine the results to yield a color in RGB notation (values between 0 and 1 for the red, green, and blue component), according to:
\begin{equation}
R = U_1 + U_2;\quad G = E + U_2;\quad B = S,
\end{equation}
where $U_1$, and $U_2$ stand for the fraction of trials displaying unstable network activity before respectively after onset of propagation, and $S$ and $E$ for the fraction of trials displaying stable background activity with respectively without persistent propagation. We obtained this scheme directly from the authors of the original study, it was only coarsely described in the original publication. Combining individual results in this way does not necessarily lead to easily interpretable colours, but using a different scheme, e.g. colouring according to the class that occurred the most often, leads to visually similar results (Fig.~\ref{fig:grid_max}).

\begin{figure}
\centering
\begin{subfigure}[t]{0.33\textwidth}
\caption{}\vspace{-1ex}
\includegraphics[width=\textwidth]{Figs/grid_l_dt_100us_gs100_mean.pdf}
\end{subfigure}%
\begin{subfigure}[t]{0.33\textwidth}
\caption{}\vspace{-1ex}
\includegraphics[width=\textwidth]{Figs/grid_nl_dt_100us_gs100_mean.pdf}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\caption{}\vspace{-1ex}
\includegraphics[width=\textwidth]{Figs/grid_nl_dt_100us_gs75_mean.pdf}
\end{subfigure}
\caption{\label{fig:grid}
\textbf{Qualitative network behaviour as a function of excitatory and inhibitory connection strength for linearly (A) and non-linearly (B, C) coupled networks.} For each combination, we initiated synchronous activity with 100 (A, B) or 75 (C) neurons and assessed the stability of the temporal evolution. Blue colouring indicates stable propagation of synchrony, red indicates unstable background activity before onset of propagation, yellow indicates unstable background activity after onset of propagation, and green colouring indicates the absence of propagation (see Methods for details).\\
Replication of Fig.~3 in \cite{Memmesheimer2012} (\url{https://doi.org/10.1371/journal.pcbi.1002384.g003})}
\end{figure}

\begin{table}
\begin{tabular}{ c  l  p{0.32\textwidth}}
  \toprule
   & Class & Definition \\
  \midrule
  $U_1$ (red) & Unstable before stimulation & $\exists t<t_\text{stim}: a_b(t) \geq 100$ \\
  $U_2$ (yellow) & Unstable after stimulation & $\exists t>t_\text{stim}: a_b(t) \geq 100$ \\
  $E$ (green) &  Stable activity, no propagation & $\forall t: a_b(t) < 100\, \wedge\, g_\text{p} < 10$\\
  $S$ (blue) & Stable activity, persistent propagation & $\forall t: a_b(t) < 100\, \wedge\, g_\text{p} \geq 10$\\
  \bottomrule
\end{tabular}
\caption{Classification of network behaviour for Fig.~\ref{fig:grid}. Here, $a_b$ refers to the background activity, the number of neurons active during a \SI{1}{\milli\second} time bin, ignoring the time steps at $t_\text{p}^k = t_\text{stim} + k\cdot\tau$, i.e. time steps that follow the synchronous stimulation by multiples of the synaptic delay. The variable $g_\text{p}$ refers to the number of successful propagations of the synchronous activity, i.e. the number of time steps $t_\text{p}^k$ where the number of active neurons exceeds the maximal background activity $a_b$.}\label{tab:color}
\end{table}
